Quantifying the Importance of Change for Understanding\\ Differences in Personal Culture

### 2024-09-13 write-ups/manuscript.tex
updates (pending case study).

\title{Quantifying the Importance of Change for Understanding\\
Differences in Personal Culture}
\author{Achim Edelmann\footnote{médialab, Science Po; equal authorship.} \and Kevin
Kiley\footnote{Department of Sociology and Anthropology, North Carolina
  State University; equal authorship.} \and Turgut
Keskintürk\footnote{Department of Sociology, Duke University} \and Isabella
Bouklas\footnote{Department of Sociology, Duke University} \and Stephen
Vaisey\footnote{Department of Sociology, Duke University}}
\date{}

@@

Recent work on this topic has converged on a general conclusion that
people do sometimes make substantial shifts in their personal culture
over time but that most elements of personal culture are characterized
by low rates of persistent change during adulthood (Kiley and Vaisey
2020; Lersch 2023). Work leading to this conclusion adopted a
``tournament of models'' approach to adjudicating different theoretical
perspectives (Lersch 2023: p.~228). In this framework, researchers
formalized different models of change over time and relied on model
selection criteria---primarily the Bayesian Information Criterion
(Raftery 1995)---to classify different questions as demonstrating either
``no change'' or ``any change.'' The size of these piles then serves as
the primary evidence for the truth of different arguments (Kiley and
Vaisey 2020; Lersch 2023; Vaisey and Kiley 2021).

However, as consensus in the overall picture emerges, researchers are
now turning toward understanding the conditions that produce persistent
change in personal culture, including age, education, and life-course
transitions; whether certain kinds of people are more likely to make
persistent changes in personal culture; and whether certain kinds of
personal culture are more likely to be characterized by persistent
change or stability. The tournament of models approach, useful for the
broad question of whether change happens, is ill-suited to addressing
these more theoretically grounded questions.

In a recent paper, Lersch (2023) takes an important step forward by
devising a Life Course Adaption Model (LCAM), a mixed-effects growth
curve model that quantifies various components of cultural difference:
early imprinting, persistent change, biographical experience, and
fluctuation. In quantifying these aspects of cultural difference, the
LCAM lets researchers address a broader range of questions than the
single question of whether a measure is characterized by stability or
change. However, because the LCAM produces so many different measures,
it is not obvious how to use these components to understand the relative
importance of within-person change, especially when change is measured
in non-linear ways.

In this paper, we build on the LCAM by outlining a single metric that
combines components of the LCAM to quantify the relative importance of
intrapersonal change and interpersonal difference for explaining
variation in a population for a variable over time. Our approach
isolates the ``systematic'' components of variance (i.e., it removes
random change and measurement error), to provide a single summary of how
much systematic variation can reasonably be attributed to people making
persistent changes over time. This quantity can then be used as a quick
diagnostic to compare across questions, groups, periods, or samples to
isolate where within-person change is more or less useful for explaining
variation. We demonstrate the utility of this quantity in two ways.

First, we use 610 items from seven panel surveys from five countries to
quantify the proportion of variance that is attributable to systematic
intrapersonal change over the study period rather than to interpersonal
differences at baseline. This allows us to generate an overall picture
of how much of the total systematic variation in personal culture change
tends to explain. These findings largely echo previous findings but
enable clearer statements about the relative importance of change across
questions, rather than binary claims about whether change ``happens'' or
specific quantification of how much change occurs.

Second, to illustrate how this quantity might be used by a researcher
with a specific topical or theoretical interest, we investigate
differences in the relative importance of intrapersonal change for
people with and without a college degree. We find that change explains a
smaller proportion of systematic variance among college-educated
respondents, suggesting that college crystallizes one's personal culture
rather than fostering an openness to new information. This offers a new
empirical basis to theorize about the role of education in influencing
personal culture.

@@

The second model summarized by Kiley and Vaisey (2020), an ``Active
Updating Model'' (AUM), posits that people continually update their
personal culture as they move through life. This model suggests people
change their personal culture as they adapt and make new meanings when
encountering new social environments, discourses, and information (Gross
2009; Swidler 2001). This model underlies, among others, theories of
cultural diffusion (Christakis and Fowler 2010), attitude alignment
(DellaPosta 2020), and polarization (Bail et al. 2018). It is also
implicit in most studies that ask whether specific experiences, changes
in social roles, or political events, affect personal culture (Gelman
and Margalit 2021; Slothuus and Bisgaard 2021; Visser and Mirabile
2004).

@@

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the
majority of items did prefer the AUM, the amount of durable change
detected on these items was small. A substantial minority of questions
(39 percent) favored the SDM, meaning they were more consistent with
zero durable change. Questions with more evidence of durable change
included salient issues like gay marriage and questions tapping
``public'' statements or behaviors such as partisan identification and
religious service attendance. There was also more evidence of durable
change among early adults (people ages 18-30) than among the rest of the
population. Overall, the researchers concluded that ``results ultimately
suggest that real, persistent attitude change is an uncommon phenomenon
among adults'\,' (Kiley and Vaisey 2020: p.~500; see also Vaisey and
Kiley 2021). This lack of durable change is consistent with other recent
findings that cohort replacement plays a larger role than period effects
in explaining differences in personal culture (Vaisey and Lizardo 2016).

@@

Drawing on these findings, Lersch (2023) challenged the SDM and AUM as a
``needless dichotomy,'' proposing the ``Life Course Adaption Model''
(LCAM) as an alternative. This model draws on the life course
perspective to model personal culture as a linear trajectory\footnote{In
  supplemental materials, Lersch models age using splines as well, with
  results generally showing the same conclusions.} over the duration of
a panel for each respondent. In doing this, Lersch rectified three
shortcomings of the AUM and SDM. First, the AUM, as described by Kiley
and Vaisey, posits that changes follow a Markov process where responses
at time \(t\) are a function of responses at time \(t-1\) but not
earlier time points. However, earlier life experiences can (directly and
indirectly) mold personal culture when transitioning to new social roles
or into new environments, even if their initial impact is delayed. For
example, childhood events might influence views on family structures
later when individuals form their own families. Therefore, the LCAM
considers influences from earlier than just \(t-1\) on responses at time
\(t\).

Second, Kiley and Vaisey's conclusions about the AUM and SDM were based
on examining three-wave panel data over four years, which might not be
intensive or extensive enough to adjudicate the two models. Lersch
evaluates the LCAM against the AUM and SDM using panel data spanning a
wider duration (from 3 to 36 years) and more waves (3 to 18), offering a
better chance to observe durable change.

Third, and most importantly, the LCAM quantifies various components of
cultural difference, allowing researchers to ask a more varied set of
questions than simply whether people change. For example, researchers
can use the LCAM framework to quantify the magnitude of change in a
measure of culture over time (i.e., how much do people change on
average?), the variance in age-based slopes over time (i.e., do people
tend to change in similar ways as they age?), or how much personal
culture tends to fluctuate around these age-based trends (i.e., how
consistent are responses over time?).

@@

Despite differences in how these researchers interpret their findings,
the empirical results are not far apart. Kiley and Vaisey (2020) found
that the majority of items they tested favored the AUM, meaning there
was evidence of persisting change among adults, a pattern consistent
with Lersch's results. Also Lersch (2023) found evidence of persisting
change on most items, but these changes were small. On average, people
changed only about .07 standard deviations over 10 years. This is
consistent with Kiley and Vaisey's finding that persisting change is
often small. Moreover, a quarter of the items Lersch studied still
preferred the SDM, suggesting that even with different assumptions about
change and more extensive data, many questions still are consistent with
a model that allows zero intrapersonal change.

Other studies are compatible with these findings as well. For example,
Brocic and Miles (2021) estimate that completing graduate degrees in
humanities, arts, and social sciences shift peoples' moral relativism
only about 0.2 standard deviations on average compared to people with no
college degree, and this is the largest effect they identify. And
studies of aggregate change show that, even on items where cohort
effects explain more variance than period effects, there is always
evidence that some people change over time (Vaisey and Lizardo 2016). In
other words, despite different interpretations, the results of previous
work are in many ways highly consonant.

@@

To this point, the debate has broadly centered on drawing conclusions
about whether people \emph{ever} change. But this is rarely the right
question since the answer is almost always ``yes.'' Over sufficient time
and in a large enough sample, researchers will likely observe some
evidence for durable change in the population. A lack of evidence for
change may be due to a poor survey question, a low resolution in
response options, or that the question simply was not asked for long
enough or to enough people. Conversely, finding evidence for some change
tells us only that it was ``not zero.'' From that alone, we learn little
about how much intrapersonal change has happened in a population.

@@

Our approach---which we formalize below---offers such a quantification.
Our measure relates two variance components that each reflect a
combination of relevant theoretical processes. The first component,
stable interpersonal differences, reflects the accumulated experiences
of people prior to entering the panel survey. Lersch (2023: 24) calls
this ``early imprinting.'' While commonly associated with experiences
during a formative period that result in settled dispositions, this
variation could also reflect experiences that happened at any time as
long as they predate the panel and consistently affect subsequent
responses. For instance, for those people entering the panel
post-retirement, this ``imprinting'' might reflect this pivotal life
transition. Consequently, this component also reflects variation in
individuals' social roles or statuses at the start of the panel that
were important in shaping their dispositions.

The second component, the amount of intrapersonal change people make
during the panel, captures persisting changes in personal culture over
time. This component captures the set of processes collectively called
``persistent change'' or ``adaption'' by Lersch and ``active updating''
by Kiley and Vaisey. Lersch attributes these changes to social triggers
such as moving into a new environment or adopting new social roles
(although he does not measure these directly). They might also reflect
the diffusion of new cultural forms across social networks, cues from
political elites or otherwise culturally influential leaders, the
emergence of issues in politics or culture, or large-scale social
shifts.

A third component, which we might call ``residual variance'' or
``fluctuation,'' accounts for the remainder of variance in peoples'
responses. These non-durable changes emerge for a variety of reasons.
For example, people might not have a clear disposition on a particular
item as it is asked. Instead, they might internalize a broad set of
considerations and construct an opinion in the context of the survey
interview, with different considerations coming to the forefront of
their cognition during each interview (Feldman and Zaller 1992;
Tourangeau, Rips, and Rasinski 2000; Zaller 1992). This variance can
also include measurement error, such as misinterpreted questions,
erroneous response selections, or responses getting coded incorrectly.
Although this third component at least partially reflects important
processes of personal culture, it does not directly touch on the ongoing
debate outlined here. Because of this, we quantify this component but
focus principally on the other two.

@@

The starting point for our quantity is Lersch's (2023) Life Course
Adaption Model, which formalizes survey responses at time \(t\) as a
function of individual-level random intercepts and slopes for survey
age. In some contexts, this model is called a mixed-effects growth curve
model. This model assumes a set of propositions about change that
reflect the theoretical debate to this point. First, consistent with the
settled disposition model, it assumes that people start the survey with
cultural differences, modeled as random intercepts for each respondent.
Second, it assumes that people change over time, taking the form of
random slopes for each respondent as a linear function of time. Third,
it assumes that people deviate around this baseline randomly over time,
reflecting ``fluctuation'' or short-term non-persistent change.
Formally, this can be written as

@@

where \(\beta_0\) is the average intercept, \(\alpha_{0i}\) is the
random intercept component for individuals, \(\beta_1\) is the average
yearly change in the outcome, \(\alpha_{1i}\) is the random slope
component for individuals, and \(\epsilon_{it}\) is the random error
term that captures transitory disturbances. The \(\alpha_{0i}\) and
\(\alpha_{1i}\) terms are also allowed to covary.

@@

Together, \(V(D)\) and \(V(C)\) reflect the total systematic variance of
the outcome that the LCAM can attribute to either differences between
people or linear intrapersonal change over the course of the panel. Our
principal measure of interest is the proportion of systematic variance
accounted for by intrapersonal change, which we calculate as

@@

Of course, the \(\omega\) measure is itself variable---most particularly
due to sampling variability---such that the composition of durable
changers and stable respondents in a particular sample might affect the
systematic components \(V(D)\) and \(V(C)\). It is thus important to
establish confidence intervals to reflect the uncertainty in the
\(\omega\). We achieve this by incorporating nonparametric bootstraps
with case resampling. In doing this, we first resample the respondents
with replacement and assign a new identifier to the set of new
\emph{respondents}, allowing us to change the composition of level-2
units in our data. Second, we estimate the LCAM in this slightly
perturbed sample, and calculate the \(\omega\). We then replicate this
procedure \(N\) times and establish confidence intervals using the
distribution of bootstrapped omega values.

@@

To demonstrate the utility of this metric, we provide two examples. In
the first, we compare the proportion of systematic variance explained by
intrapersonal change across a broad array of questions.

We use data from seven nationally representative panel surveys from
Australia, Germany, Great Britain, Switzerland, and the United States
(summarized in Table 1), combining all the data files used in previous
work (Kiley and Vaisey 2020; Lersch 2023).\footnote{For more information
  on these data sources, see (Goebel et al. 2019; Income Dynamics 2013;
  Smith et al. 2022; Summerfield et al. 2011; Taylor 1996; University of
  Essex and Research 2019; Voorpostel et al. 2016).} These studies cover
a long period of time (the range of surveys spans from 1968 to 2021),
with 609 personal culture items capturing attitudes, beliefs, values,
self-assessments, self-descriptions, and behaviors (Alwin 2007). We
restricted the sample such that individuals between the ages of 18 and
79 are included without further elimination, and in all surveys, we used
all possible cases for which respondents provided responses. In the end,
the analyses that follow rely on a cross-national sample with a
cross-domain set of items to capture a broad range of individual
personal culture. Supplemental Materials A documents the list of all
variables used in the upcoming analyses.

\begin{table}[H]
\centering
\caption{\label{tab:unnamed-chunk-1}The Description of the Data Sources}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}[t]{llcc}
\toprule
Country & Survey & Period & Outcomes\\
\midrule
Australia & Household, Income and Labor Dynamics in Australia (HILDA) & 2001-2021 & 30\\
Germany & Socio-Economic Panel Study (SOEP) & 1984-2020 & 122\\
Great Britian & British Household Panel Survey (BHPS) & 1991-2008 & 83\\
Great Britian & Understanding Society/UK Household Longitudinal Study (UKHLS) & 2009-2020 & 66\\
Switzerland & Swiss Household Panel (SHP) & 1999-2019 & 77\\
\addlinespace
United States & General Social Survey (GSS) & 2006-2012 & 183\\
United States & Panel Study of Income Dynamics (PSID) & 1968-2019 & 49\\
\bottomrule
\end{tabular}}
\end{table}

\subsection{Analysis}\label{analysis}

As a first step in our analysis, we fit this LCAM to each of the 609
measures of personal culture outlined above. We then derive \(\omega\)
for each of these items. As a second step, we use linear regression to
model \(\omega\) as a function of features of the question and panel.
This allows us to explore how these features are associated with the
degree of intrapersonal change. These covariates include the number of
response options (measured as 2, 3-5, and more than 5), the specific
survey, the log of the number of participants, the number of waves
observed, the time period (i.e., decade) the question was first asked,
and the total duration of time the question was asked.

@@

Figure 1 plots the proportion of systematic variance attributable to
interpersonal differences and intrapersonal change for each of the 609
questions, plotted separately by panel. To broadly summarize the
results, all panels show a similar range of the proportion of systematic
variance attributable to intrapersonal change. Across panels, the median
value of \(\omega\) is 0.094 (mean of 0.098), with an interquartile
range of 0.053 to 0.139. All panels include questions where \(\omega\)
is essentially 0, and all panels include questions with \(\omega\)
values greater than 0.20. The maximum \(\omega\) value is 0.26.

\begin{figure}[ht]
\begin{center}
\caption{Proportions of Systematic Variance in Personal Culture}

@@

\end{center}
\footnotesize{\textit{Notes:} The figure shows $\omega$ and 1 - $\omega$ as the proportion of systematic variance attributable to intrapersonal change and interpersonal differences. See Supplemental Materials A for the full set of item values.}
\end{figure}

Across all questions, interpersonal differences account for a much
larger share of the systematic variance in responses than intrapersonal
change. Again, this is to be expected. Interpersonal differences capture
not just pre-adult socialization, but all accumulated experiences up to
the start of the panel that might influence personal culture.

To the extent that there are differences across the panels, the PSID has
the highest \(\omega\) values with mean .120 and median .135. While we
cannot disentangle features of the sample from features of the questions
asked to each sample, the specific samples for many PSID questions have
lower average ages than those from other panels. To the extent that
younger respondents might be more likely to make durable changes of
opinion, these higher estimates of intrapersonal change might reflect
the distinct age profile of respondents in this sample. At the other
end, the GSS has the lowest range of \(\omega\) with mean .073 and
median .069. This potentially reflects the fact that the GSS observes
people for a shorter duration, on average, than the other panels. If,
consistent with life course adaption theories, people are more likely to
make significant cultural changes the longer we observe them, then
duration likely affects the range of \(\omega\) (a point we explore more
below). However, the GSS results are still consistent with results from
the other panels.

While there are some differences between panels, these differences are
small compared to the differences within panels. For about 6 percent of
items, \(\omega\) is greater than 0.20. These questions tend to ask
about objectively changing external referents (e.g., confidence in
specific government leaders or political parties), life satisfaction, or
current financial position. At the other end, questions about religious
identification, views on gender roles, and support for civil liberties
tend to have very low estimates of intrapersonal change.

In contrast to the tournament of models approach, quantifying change
this way allows us to explore variation in the relative importance of
intrapersonal change across questions that all show evidence of change.
For example, Kiley and Vaisey (2020) found that confidence in the press
and confidence in religious leaders were both characterized by active
updating. Our results show that intrapersonal change is much more
important for explaining variance in confidence in the press (0.164)
than confidence in religion (0.049), even though both are updating.

Appendix A shows the distribution of \(V(D)\) and \(V(C)\) across
panels, and Supplemental Materials A presents the estimated proportion
of variance attributable to interpersonal differences and intrapersonal
change, the estimated values of \(\omega\), and the proportion of
residual variance for each question. Interpersonal differences are
almost always the largest component of the total variance and tend to
account for between 55 and 70 percent of total variance, while
intrapersonal change is always the smallest, typically accounting for
between 3 and 8 percent of total variance. Residual variance tends to
account for between 22 and 37 percent of variance, though on several
questions residual variance is greater than 50 percent. This might
indicate survey items with low reliability or ones that capture
genuinely rapid fluctuations.

>>>>

\title{Quantifying the Importance of Change for Understanding\\
Differences in Personal Culture}
\author{Achim Edelmann\footnote{médialab, Science Po; equal authorship.} \and Kevin
Kiley\footnote{Department of Sociology and Anthropology, North Carolina
  State University; equal authorship.} \and Turgut
Keskintürk\footnote{Department of Sociology, Duke University} \and Isabella
Bouklas\footnote{Department of Sociology, Duke University} \and Stephen
Vaisey\footnote{Department of Sociology, Duke University}}
\date{September 2024}

@@

Recent work on this topic has converged on a general conclusion that
people do sometimes make substantial shifts in their personal culture
over time but that most elements of personal culture are characterized
by low rates of persistent change during adulthood (Kiley and Vaisey
2020; Lersch 2023). Work leading to this conclusion adopted a
``tournament of models'' approach to adjudicating different theoretical
perspectives (Lersch 2023:228). In this framework, researchers
formalized different models of change over time and relied on model
selection criteria---primarily the Bayesian Information Criterion
(Raftery 1995)---to classify different questions as demonstrating either
``no change'' or ``any change.'' The size of these piles then serves as
the primary evidence for the truth of different arguments (Kiley and
Vaisey 2020; Lersch 2023; Vaisey and Kiley 2021).

However, as consensus emerges, researchers are now turning toward
understanding the conditions that produce persistent change in personal
culture, including age, education, and life-course transitions; whether
certain kinds of people are more likely to make persistent changes in
personal culture; and whether certain kinds of personal culture are more
likely to be characterized by persistent change or stability. The
tournament of models approach, useful for the broad question of whether
change happens, is ill-suited to addressing these more theoretically
grounded questions.

In a recent paper, Lersch (2023) takes an important step forward by
devising a Life Course Adaption Model (LCAM), a mixed-effects growth
curve model that quantifies various components of cultural difference:
early imprinting, persistent change, biographical experience, and
fluctuation. In quantifying these aspects of cultural difference, the
LCAM lets researchers address a broader range of questions than the
single question of whether a measure is characterized by stability or
change. However, because the LCAM produces so many different measures,
it is not obvious how to use these components to understand the relative
importance of within-person change across items, groups, or time
periods, especially when cultural change is measured in non-linear ways.

In this paper, we build on the LCAM by outlining a single metric that
combines components of the LCAM to quantify the relative importance of
intrapersonal change and interpersonal difference for explaining
variation in a population for a variable over time. Our approach
isolates the ``systematic'' components of variance (i.e., it removes
random change and measurement error), to provide a single summary of how
much systematic variation can reasonably be attributed to people making
persistent changes over time. This quantity---what we call
\(\omega\)---can then be used as a diagnostic to compare across
questions, groups, or periods to isolate where within-person change is
more or less useful for explaining aggregate variation. We demonstrate
the utility of this quantity in two ways.

First, we use 610 items from seven panel surveys from five countries to
quantify the proportion of variance that is attributable to systematic
intrapersonal change over the study period rather than to interpersonal
differences at baseline. This allows us to generate an overall picture
of how much of the total systematic variation in personal culture change
tends to explain. These findings largely echo previous findings, but
enable clearer statements about the relative importance of change across
questions, rather than binary claims about whether change ``happens.''

Second, to illustrate how this quantity might be used by a researcher
with a specific theoretical interest, we investigate differences in the
relative importance of intrapersonal change for people with and without
a college degree, using several political preferences in the Swiss
Household Panel as a case study. We find that change explains a smaller
proportion of systematic variance among college-educated respondents,
suggesting that college crystallizes one's personal culture rather than
fostering an openness to new information. This offers a new empirical
basis to theorize about the role of education in influencing personal
culture, while at the same time showcasing a simple use of our proposed
\(\omega\) measure in a substantive theoretical context.

@@

The second model summarized by Kiley and Vaisey (2020), an ``Active
Updating Model'' (AUM), posits that people continually update their
personal culture as they move through life. This model suggests people
change their personal culture as they adapt and make new meanings when
encountering new social environments, discourses, and information (Gross
2009; Swidler 2001). This model underlies, among others, theories of
cultural diffusion (Christakis and Fowler 2010), attitude alignment
(DellaPosta 2020), and polarization (Bail et al. 2018). It is also
implicit in most studies that ask whether specific experiences, changes
in social roles, or political events, affect personal culture (Gelman
and Margalit 2021; Slothuus and Bisgaard 2021).

@@

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the
majority of items did prefer the AUM, the amount of durable change
detected on these items was small. A substantial minority of questions
(39 percent) favored the SDM, meaning they were more consistent with
zero durable change. Questions with more evidence of durable change
included salient issues like gay marriage and questions tapping
``public'' statements or behaviors such as partisan identification and
religious service attendance. There was also more evidence of durable
change among early adults (people ages 18-30) than among the rest of the
population. Overall, the researchers concluded that ``results ultimately
suggest that real, persistent attitude change is an uncommon phenomenon
among adults'' (Kiley and Vaisey 2020:500). This lack of durable change
is consistent with other recent findings that cohort replacement plays a
larger role than period effects in explaining differences in personal
culture (Vaisey and Lizardo 2016).

@@

Drawing on these findings, Lersch (2023) challenged the SDM and AUM as a
``needless dichotomy,'' proposing the ``Life Course Adaption Model''
(LCAM) as an alternative. This model draws on the life course
perspective to model personal culture as a linear trajectory\footnote{In
  supplemental materials, Lersch uses splines as well, with results
  generally showing the same conclusions.} over the duration of a panel
for each respondent. In doing this, Lersch rectified three shortcomings
of the AUM and SDM. First, the AUM, as described by Kiley and Vaisey,
posits that changes follow a Markov process where responses at time
\(t\) are a function of responses at time \(t-1\), but not earlier time
points. However, earlier life experiences can (directly and indirectly)
mold personal culture when transitioning to new social roles or into new
environments, even if their initial impact is delayed. For example,
childhood events might influence views on family structures later when
individuals form their own families. Therefore, the LCAM considers
earlier influences on responses at time \(t\).

Second, Kiley and Vaisey's conclusions about the AUM and SDM were based
on examining three-wave panel data over four years, which might not be
intensive or extensive enough to adjudicate the two models. Lersch
evaluates the LCAM against the AUM and SDM using panel data spanning a
wider duration (from 3 to 36 years) and more waves at the respondent
level (3 to 18), offering a better chance to observe durable change in a
longer time frame.

Third, and most importantly, the LCAM quantifies various components of
cultural difference, allowing researchers to ask a more varied set of
questions than simply whether people change. For example, researchers
can use the LCAM framework to quantify the magnitude of change in a
measure of culture over time (how much do people change on average?),
the variance in age-based slopes over time (do people tend to change in
similar ways as they age?), or how much personal culture tends to
fluctuate around these age-based trends (how consistent are responses
over time?).

@@

Despite differences in how these researchers interpret their findings,
the empirical results are not far apart. Kiley and Vaisey (2020) found
that the majority of items they tested favored the AUM, meaning there
was evidence of persisting change among adults, a pattern consistent
with Lersch's results. Also, Lersch (2023) found evidence of persisting
change on most items, but these changes were relatively small. On
average, people changed only about .07 standard deviations over 10
years. This is consistent with Kiley and Vaisey's finding that
persisting change is often small in magnitude. Moreover, a quarter of
the items Lersch studied still preferred the SDM, suggesting that even
with different assumptions about change and more extensive data, many
questions still are consistent with a model that allows zero
intrapersonal change.

Other studies are compatible with these findings as well. For example,
Brocic and Miles (2021) estimate that completing graduate degrees in
humanities, arts, and social sciences shift peoples' moral relativism
only about .20 standard deviations on average compared to people with no
college degree, and this is the largest effect they identify. And
studies of aggregate change show that, even on items where cohort
effects explain more variance than period effects, there is always
evidence that some people change over time (Vaisey and Lizardo 2016). In
other words, despite different interpretations, the results of previous
work are in many ways highly consonant.

@@

To this point, the debate has broadly centered on drawing conclusions
about whether people \emph{ever} change. But this is rarely the right
question since the answer is almost always ``yes.'' Over sufficient time
and in a large enough sample, researchers will likely observe some
evidence for change in the population. A lack of evidence for change may
be due to a poor survey question, a low resolution in response options,
or that the question simply was not asked for long enough or to enough
people. Conversely, finding evidence for some change tells us only that
it was ``not zero.'' From that alone, we learn little about how much
intrapersonal change has happened in a population.

@@

Our approach---which we formalize below---offers such a quantification.
Our measure relates two variance components that each reflect a
combination of relevant theoretical processes. The first component,
stable interpersonal differences, reflects the accumulated experiences
of people prior to entering the panel survey. Lersch (2023:24) calls
this ``early imprinting.'' While commonly associated with experiences
during a formative period that result in settled dispositions, this
variation could also reflect experiences that happened at any time as
long as they predate the panel and consistently affect subsequent
responses. For instance, for those people entering the panel
post-retirement, this ``imprinting'' might reflect this pivotal life
transition and its subsequent effects observed in the panel window.
Consequently, this component also reflects variation in individuals'
social roles or statuses at the start of the panel that were important
in shaping their dispositions.

The second component, the amount of intrapersonal change people make
during the panel, captures persisting changes in personal culture over
time. This component captures the set of processes collectively called
``persistent change'' or ``adaption'' by Lersch (2023) and ``active
updating'' by Kiley and Vaisey (2020). Lersch (2023) attributes these
changes to social triggers such as moving into a new environment or
adopting new social roles (although he does not measure these directly).
They might also reflect the diffusion of new cultural forms or ideas
across social networks, cues from political elites or otherwise
culturally influential leaders, the emergence of issues in politics or
culture, or large-scale social shifts.

A third component, which we might call ``residual variance'' or
``fluctuation,'' accounts for the remainder of variance in peoples'
responses. These non-durable changes emerge for a variety of reasons.
For example, people might not have a clear disposition on a particular
item as it is asked. Instead, they might internalize a broad set of
considerations and construct an opinion in the context of the survey
interview, with different considerations coming to the forefront of
their cognition during each interview (Feldman and Zaller 1992;
Tourangeau, Rips, and Rasinski 2000; Zaller 1992). This variance can
also include measurement error, such as misinterpreted questions,
erroneous response selections, or responses getting coded incorrectly.
Although this third component at least partially reflects important
processes of personal culture, it does not directly touch on the ongoing
debate outlined here. Hence, we quantify this component but focus
principally on the other two.

@@

The starting point for our quantity is Lersch's (2023) Life Course
Adaption Model, which formalizes survey responses at time \(t\) as a
function of individual-level random intercepts and slopes for survey
age. In some contexts, this model is called a mixed-effects growth curve
model. This model assumes a set of propositions about change that
reflect the theoretical debate to this point. First, consistent with the
SDM, it assumes that people start the survey with cultural differences,
modeled as random intercepts for each respondent. Second, it assumes
that people change over time, taking the form of random slopes for each
respondent as a linear function of time. Third, it assumes that people
deviate around this baseline randomly over time, reflecting
``fluctuation'' or short-term non-persistent change. Formally, this can
be written as

@@

where \(\beta_0\) is the average intercept, \(\alpha_{0i}\) is the
random intercept component for individuals, \(\beta_1\) is the average
yearly change in the outcome, \(\alpha_{1i}\) is the random slope
component for individuals, and \(\epsilon_{it}\) is the random error
term that captures fluctuations. The \(\alpha_{0i}\) and \(\alpha_{1i}\)
terms are also allowed to covary.

@@

Together, \(V(D)\) and \(V(C)\) reflect the total systematic variance of
the outcome that the LCAM can attribute to either differences between
people or linear intrapersonal change\footnote{The decomposition of
  variance is agnostic to this choice. One advantage is that the
  modeling assumptions can be relaxed to accommodate non-linear terms or
  splines, but the decomposition will follow the same procedure.} over
the course of the panel. Our principal measure of interest is the
proportion of systematic variance accounted for by intrapersonal change,
which we calculate as

@@

Of course, the \(\omega\) measure is itself variable---most particularly
due to sampling variability---such that the composition of durable
changers and stable respondents in a particular sample might affect the
systematic components \(V(D)\) and \(V(C)\). It is thus important to
establish confidence intervals to reflect the uncertainty in the
\(\omega\). We achieve this by incorporating nonparametric bootstraps
with case resampling. In doing this, we first resample the respondents
with replacement and assign a new identifier to the set of new
\emph{respondents}, allowing us to change the composition of level-2
units in our data. Second, we estimate the LCAM in this slightly
perturbed sample, and calculate the \(\omega\). We then replicate this
procedure \(N\) times and establish confidence intervals around our
measure.

@@

To demonstrate the utility of \(\omega\), we provide two examples. In
the first, we compare the proportion of systematic variance explained by
intrapersonal change across a broad array of questions.

\subsection{Data and Analysis}\label{data-and-analysis}

We use data from seven nationally representative panel surveys from
Australia, Germany, Great Britain, Switzerland, and the United States
(summarized in Table 1), combining all the data files used in previous
work (Kiley and Vaisey 2020; Lersch 2023).\footnote{For more information
  on these data sources, see the relevant panel files (Goebel et al.
  2019; Income Dynamics 2013; Smith et al. 2022; Summerfield et al.
  2011; Taylor 1996; University of Essex and Research 2019; Voorpostel
  et al. 2016).} These studies cover a long period of time (the range of
surveys spans from 1968 to 2021), with 610 personal culture items
capturing attitudes, beliefs, values, self-assessments,
self-descriptions, and behaviors (Alwin 2007). We restricted the sample
such that individuals between the ages of 18 and 79 are included without
further elimination, and in all surveys, we used all possible cases for
which respondents provided responses. In the end, the analyses that
follow rely on a cross-national sample with a cross-domain set of items
to capture a broad range of individual personal culture. We document the
list of all variables used in the upcoming analyses in the Appendix.

\begin{table}[H]
\centering
\caption{\label{tab:unnamed-chunk-1}The Description of the Data Sources}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}[t]{llcc}
\toprule
Country & Survey & Period & Outcomes\\
\midrule
Australia & Household, Income and Labor Dynamics in Australia (HILDA) & 2001-2021 & 30\\
Germany & Socio-Economic Panel Study (SOEP) & 1984-2020 & 122\\
Great Britian & British Household Panel Survey (BHPS) & 1991-2008 & 83\\
Great Britian & Understanding Society/UK Household Longitudinal Study (UKHLS) & 2009-2020 & 66\\
Switzerland & Swiss Household Panel (SHP) & 1999-2019 & 77\\
United States & General Social Survey (GSS) & 2006-2012 & 183\\
United States & Panel Study of Income Dynamics (PSID) & 1968-2019 & 49\\
\bottomrule
\end{tabular}}
\end{table}

As a first step in our analysis, we fit this LCAM to each of the 610
measures of personal culture outlined above. We then derive \(\omega\)
for each of these items using the procedure above.

@@

Figure 1 plots the proportion of systematic variance attributable to
interpersonal differences and intrapersonal change for each of the 610
questions, plotted separately by panel survey. To broadly summarize the
results, all panels show a similar range of the proportion of systematic
variance attributable to intrapersonal change. Across panels, the median
value of \(\omega\) is .053 (mean of .056), with an interquartile range
of .032 to .077. All panels include questions where \(\omega\) is
essentially 0, and some panels include questions with \(\omega\) values
greater than .20. The maximum \(\omega\) value is .23.

\begin{figure}[!p]
\begin{center}

@@

\end{center}
\caption*{Figure 1: Proportions of Systematic Variance in Personal Culture}
\footnotesize{\textit{Notes:} The figure shows $\omega$ and 1 - $\omega$ as the proportion of systematic variance attributable to intrapersonal change and interpersonal differences. See the Appendix for the full set of item values.}
\end{figure}

Across all questions and all surveys, interpersonal differences account
for a much larger share of the systematic variance in responses than
intrapersonal change. Again, this is to be expected. Interpersonal
differences capture not just pre-adult socialization, but all
accumulated experiences up to the start of the panel that might
influence personal culture.

To the extent that there are differences across the panels, the PSID has
the highest \(\omega\) values with mean .068 and median .076. While we
cannot disentangle features of the sample from features of the questions
asked to each sample, the specific samples for many PSID questions have
lower average ages than those from other panels. To the extent that
younger respondents might be more likely to make durable changes of
opinion, these higher estimates of intrapersonal change might reflect
the distinct age profile of respondents in this sample. At the other
end, the SHP has the lowest range with mean .042 and median .037,
followed by GSS, with mean .044 and median .043.

While there are some differences between panels, these differences are
small compared to the differences within panels. For about 9 percent of
items, \(\omega\) is greater than .10. These questions tend to ask about
objectively changing external referents (e.g., confidence in specific
government leaders or political parties), life satisfaction, or current
financial position. At the other end, questions about religious
identification, views on gender roles, and support for civil liberties
tend to have very low estimates of intrapersonal change, suggesting that
core preferences might be less variable.

In contrast to the tournament of models approach, quantifying change
this way allows us to explore variation in the relative importance of
intrapersonal change across questions that all show evidence of change.
For example, Kiley and Vaisey (2020) found that confidence in the press
and confidence in religious leaders were both characterized by active
updating. Our results show that intrapersonal change is much more
important for explaining variance in confidence in the press (0.091)
than confidence in religion (0.051), even though in both cases people
are updating their beliefs.

In the Appendix, we show the distribution of \(V(D)\), \(V(C)\) and the
residual variance \(V(E)\) for each question across panels.
Interpersonal differences are almost always the largest component of the
total variance and tend to account for between 50 and 66 percent of
total variance, while intrapersonal change is always the smallest,
typically accounting for between 2 and 5 percent of total variance.
Residual variance tends to account for between 30 and 46 percent of
variance, though on several questions residual variance is greater than
50 percent. This might indicate survey items with low reliability or
ones that capture genuinely rapid fluctuations.
